Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic

### 2021-08-25 write-up/covid-writeup.tex
paper almost done
Paper done outside final checks (need to carefully explain why analysis isnt great)

 %\subsubsection{Uncertain sensitivity and specificity}
 %
 %For prevalence there is some
 %
 %For
 %
 %For seroprevalence, \cite{Bendavid2020} report
 %
 %
 %
 %, which leads to a range for $M$ of $(2.21, 2.38)$.  Therefore, the impact on bias of $\hat R_t$ appears moderate.  If $FP = 0.05$ and $FN = 0.005$, then the relative sampling rate increases to $M=3.04$.  This simple tool helps us understand how sensitivity and specificity impact biases and  therefore how much we should trust conclusions based on these assumptions.
 %
 %
 \subsubsection{Asymptotics}

@@

 Unfortunately, the first term of the doubly robust estimator cannot be computed as covariate information is not collected on the entire population.  Here, we again use the probability samples and consider the doubly-robust estimator
 $$
 \frac{1}{\sum_{j=1}^N \tilde I_j \tilde W_j} \sum_{j=1}^N  \tilde I_j \tilde W_j \hat \mu (x_j) + \frac{1}{\sum_{j=1}^N I_j w (x_j)} \sum_{j=1}^N I_j w(x_j) \left( \frac{Y_j - FP}{1 - FP - FN} - \mu(x_j) \right).
 $$
 where $\tilde I_j$ indicates an individual is included in the random sample and $\tilde W_j$ is the associated weight for that unit. Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:dr} shows that the IPW estimator is  consistent and calculates the estimator's variance. See Appendix~\ref{app:asympderivations} for additional details.
 \textcolor{red}{clarify what you are predicting as we go from no time to longitudinal setting. WE can't assume the survey at the time so we need the kernel again here.}

@@

 To make the approach concrete, Figure~\ref{fig:testinglik_mainpaper} presents the testing propensity for non-hispanic, white males across age strata and fever status using the second approach.  We note here that the ratio of testing propensity within strata across fever status is time-varying.  The ratio starting nine times more likely in early April and dropping to twice as likely by end of 2020.

@@

 Here, we construct active infection rate estimates using the unweighted, inverse-probability weighted, model-based, and doubly robust estimators.  Figure~\ref{fig:air} presents the results. The inverse-probability weighting methods do not significantly reduce bias compared to the unweighted estimates.  We conjecture this is due to the limited availability of symptom and other important covariates that may impact the likelihood an individual will get  tested. Our imputation strategies use available data to try and alleviate this issue; however, this suggests better data collection in the future may strongly improve the IPW estimates as seen in Section~\ref{section:ipwapproach} when analyzing the statewide sample from April 25th--29th.

@@

 \begin{figure}
 \centering
 \begin{subfigure}{.5\textwidth}
   \centering
   \includegraphics[width=.9\linewidth]{../figs/tv_air.png}
   \caption{Active Infection Rate Estimates}
   \label{fig:air}
 \end{subfigure}%
 \begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{../figs/tv_air_ratio.png}
  \caption{Ratio estimates}
  \label{fig:air_ratio}
 \end{subfigure}
 \caption{Time-varying estimates based on unweighted, model-based, IPW and doubly robust methods.}
 \label{fig:tvestimates}
 \end{figure}

 We also construct ratio estimates as discussed in Section~\ref{section:rates}. Figure~\ref{fig:air_ratio} presents the results.  Unweighted, IPW, and doubly robust estimates appear similar.  Recall the severe bias during the growth phase due to selection bias dicussed in Section~\ref{section:rates} may imply the observed rates are  overestimated between October 2020 and January 2021.  The model-based methods may be as well.

@@

 \subsection*{High quality information on prevalence.}

 Governments are pushing for increased testing capacity and robust contact tracing.  Contact tracing is incredibly useful for identifying carriers early and preventing spread of the disease.  We argue that understanding active infections is also key.  We know that the probability of an outbreak is a function of current prevalence and network connectivity.  Epidemic critical thresholds have been derived for many infectious disease models~\citep{Pastor2001,Newman2002,Parshani2010}.  Knowing prevalence will help governments determine long-term community-level risk and allow for more targeted interventions -- shutting down only certain locations when necessary.

@@

 While we are eternally grateful to Indiana for the release, symptom information was not made available.  This severely limits our ability to estimate correct selection propensity models.  What is required in the future is a more detailed effort to collect relevant measures of an individual's past history that may impact their likelihood of seeking testing and being a positive COVID-19 case.  Such data collection and dissemination is critical to address selection bias.

 A valid criticism of the proposed approach is that recording covariates on every tested individual is too time-consuming and costly, i.e., the data collection does not scale well.  We argue that the list of relevant factors is not too long.  For example, symptom status and COVID-19 contact are clearly relevant factors in testing selection.  Future work may consider how to include/exclude various factors over time that are not significant to limit citizen burden in reporting.  Moreover, equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample; however, this is often not feasible for state and local governments where rapid sample collection is prioritized over additional collection of covariate information.  We can easily extend~\eqref{eq:auxinfoprob} by randomly sampling a subset of the nonprobability sample on which to collect the covariate information, yielding a new pseudo-likelihood
 \begin{equation*}
 % \label{eq:auxinfoprob2}
 \sum_{j=1}^N I_j I_j^{(2)} W_j^{(2)} \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
 \end{equation*}
 Here, $I_j^{(2)}$ is an indicator that the individual is in the probability subsample and $W_j^{(2)}$ is the probability of inclusion.  Estimation proceeds by maximizing the pseudo-likelihood which now depends on random sampling in both the first and second term.  This would allow for rapid testing and minimizes cost of covariate collection.

>>>>

 %\subsubsection{Uncertain sensitivity and specificity}
 %
 %For prevalence there is some
 %
 %For
 %
 %For seroprevalence, \cite{Bendavid2020} report
 %
 %
 %
 %, which leads to a range for $M$ of $(2.21, 2.38)$.  Therefore, the impact on bias of $\hat R_t$ appears moderate.  If $FP = 0.05$ and $FN = 0.005$, then the relative sampling rate increases to $M=3.04$.  This simple tool helps us understand how sensitivity and specificity impact biases and  therefore how much we should trust conclusions based on these assumptions.
 %
 %
 \subsubsection{Asymptotics}
 \label{section:asymptotics}

@@

 Unfortunately, the first term of the doubly robust estimator cannot be computed as covariate information is not collected on the entire population.  Here, we suppose the same asymptotic regime as in Section~\ref{section:asymptotics} and therefore use the probability samples and consider the doubly-robust estimator at time~$t$
 \begin{align*}
 &\frac{1}{\sum_{l=1}^L K_{t,t_l^\prime} \sum_{j=1}^N \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime}} \sum_{l=1}^L K_{t,t_l^\prime} \sum_{j=1}^N  \tilde I_{j,t_l^\prime} \tilde W_{j,t_l^\prime} \hat \mu (X_{j,t^\prime}) \\
 + &\quad \frac{1}{\sum_{j=1}^N I_j w (X_{j,t})} \sum_{j=1}^N I_j w(X_{j,t}) \left( \frac{Y_{j,t} - FP}{1 - FP - FN} - \mu_t(X_{j,t}) \right).
 \end{align*}
 where $\tilde I_{j,t_l^\prime}$ indicates an individual is included in the random sample at time~$t_l^\prime$ and $\tilde W_{j,t_l^\prime}$ is the associated weight for that unit. Assuming the selection propensity model is correctly specified, then under the design of the probability sample and design of the subsample from the nonprobability sample, Lemma~\ref{lemma:dr} shows that the doubly robust estimator is consistent and calculates the estimator's variance. See Appendix~\ref{app:asympderivations} for additional details.

@@

 To make the approach concrete, Figure~\ref{fig:testinglik_mainpaper} presents the testing propensity for non-hispanic, white males across age strata and fever status using the second approach.  We note here that the ratio of testing propensity within strata across fever status is time-varying.  The ratio starting 50 times more likely in early April and dropping to ten times as likely by end of 2020.

@@

 Here, we construct active infection rate estimates using the unweighted, inverse-probability weighted, model-based, and doubly robust estimators.  Figure~\ref{fig:tvestimates} presents the results. The inverse-probability weighting methods do not significantly reduce bias compared to the unweighted estimates. We conjecture the poor performance is due to the limited availability of symptom and other important covariate information to estimate strata-specific infection rates. This suggests better data collection in the future may strongly improve the IPW estimates as seen in Section~\ref{section:ipwapproach} when analyzing the statewide sample from April 25th--29th.

@@

  % We also construct ratio estimates as discussed in Section~\ref{section:rates}. Figure~\ref{fig:air_ratio} presents the results.  Unweighted, IPW, and doubly robust estimates appear similar.  Recall the severe bias during the growth phase due to selection bias dicussed in Section~\ref{section:rates} may imply the observed rates are  overestimated between October 2020 and January 2021.  The model-based methods may be as well.

  \begin{figure}[!th]
 % \centering
 % \begin{subfigure}{.5\textwidth}
   \centering
   \includegraphics[width=.6\linewidth]{../figs/tv_air.png}
   % \caption{Active Infection Rate Estimates}
   % \label{fig:air}
 % \end{subfigure}%
 % \begin{subfigure}{.5\textwidth}
 %  \centering
 %  \includegraphics[width=.9\linewidth]{../figs/tv_air_ratio.png}
 %  \caption{Ratio estimates}
 %  \label{fig:air_ratio}
 % \end{subfigure}
 \caption{Time-varying active infectino rate estimates based on unweighted, model-based, IPW and doubly robust methods.}
 \label{fig:tvestimates}
 \end{figure}

@@

 % \subsection*{High quality information on prevalence.}

 % Governments are pushing for increased testing capacity and robust contact tracing.  Contact tracing is incredibly useful for identifying carriers early and preventing spread of the disease.  We argue that understanding active infections is also key.  We know that the probability of an outbreak is a function of current prevalence and network connectivity.  Epidemic critical thresholds have been derived for many infectious disease models~\citep{Pastor2001,Newman2002,Parshani2010}.  Knowing prevalence will help governments determine long-term community-level risk and allow for more targeted interventions -- shutting down only certain locations when necessary.

@@

 While we are eternally grateful to Indiana for the release, symptom information was not made available.  This severely limits our ability to estimate correct selection propensity models and strata-specific infection rates.  A more sustained effort to collect relevant measures of an individual's past history that may impact their likelihood of seeking testing and being COVID-19 positive is required.  Such data collection and dissemination is critical to address selection bias.

 A valid criticism of the proposed approach is that recording covariates on every tested individual is too time-consuming and costly, i.e., the data collection does not scale well.  We argue that the list of relevant factors is not too long.  For example, symptom status and COVID-19 contact are clearly relevant factors in testing selection.  Future work may consider how to include/exclude various factors over time that are not significant to limit citizen burden in reporting.  Moreover, equation~\eqref{eq:auxinfoprob} presupposes covariate information is collected for all individuals in the nonprobability sample; however, this is often not feasible for state and local governments where rapid sample collection is prioritized over additional collection of covariate information.  We can easily extend~\eqref{eq:auxinfoprob} by randomly sampling a subset of the nonprobability sample on which to collect the covariate information.
 %yielding a new pseudo-likelihood
 % \begin{equation*}
 % % \label{eq:auxinfoprob2}
 % \sum_{j=1}^N I_j I_j^{(2)} W_j^{(2)} \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right)  + \sum_{i=1}^N \tilde I_j \tilde W_j \log ( 1 - \pi (X_j; \theta)).
 % \end{equation*}
 % Here, $I_j^{(2)}$ is an indicator that the individual is in the probability subsample and $W_j^{(2)}$ is the probability of inclusion. Estimation proceeds by maximizing the pseudo-likelihood which now depends on random sampling in both the first and second term.
 This would allow for rapid testing and minimizes cost of covariate collection.
